---
title: "LUCID: Mortality analysis"
output:
  html_document:
    toc: true
    toc_float: true
---
    
```{r setup, include=FALSE}
lib_paths <- .libPaths()
lib_paths <- c("//win.ad.jhu.edu/Users$/HOME/R/win-library/4.0", lib_paths)
.libPaths(lib_paths)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center")
```





# Load packages and read data

Load packages and utility functions. Read in clean data.

```{r message=FALSE}
library(SummarizedExperiment)
library(matrixStats)
library(readxl)
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(purrr)
library(stringr)
library(lubridate)
library(limma)
library(splines)
library(glue)
library(caret)
library(glmnet)
library(randomForest)

source("functions_utils.R")
source("functions_limma_analysis.R")
source("functions_ml.R")

load("../data/summ_exps_imputed_ruv.rda")
```

Read in Eugene's annotations to identify drug metabolites.

```{r}
metab_annots <- read_excel("../data/Broad data--redundant metabolites_02032022.xlsx", sheet = "Sheet1", range = "A1:I665", col_names = TRUE)
colnames(metab_annots) <- fix_names(colnames(metab_annots))
metab_annots <- metab_annots %>% select(-`...6`)

## Make sure that there is only one way that "medication" is spelled
metab_annots %>% dplyr::count(comment)

drug_metabs <- metab_annots %>%
    filter(comment=="medication") %>%
    pull(metabolite)
```





# Remove internal standards and count metabolites pre-filtering

```{r}
se_c8_main <- remove_internal_standards(se_c8_main)
se_hilic_pos_main <- remove_internal_standards(se_hilic_pos_main)
se_hilic_neg_main <- remove_internal_standards(se_hilic_neg_main)
```

```{r}
se_list <- list(
    c8 = se_c8_main,
    hilic_pos = se_hilic_pos_main,
    hilic_neg = se_hilic_neg_main
)
se <- merge_ses_across_injections(se_list)
nrow(se)
rowData(se) %>% as.data.frame() %>% count(is.na(hmdb_id))
```





# Explore distribution of days to death

Explore distribution of days from dialysis initiation to death: in Frenova and Canada separately and combined.

```{r}
cd_c8 <- colData(se_c8_main) %>% as.data.frame()
cd_c8_subs <- cd_c8 %>% filter(country=="Canada" | frenova, event_name == "Baseline")
ggplot(cd_c8_subs, aes(x = daysdeath, color = country)) +
    geom_density()
cd_c8_subs %>%
    group_by(country) %>%
    summarize(mean_dd = mean(daysdeath, na.rm = TRUE), sd_dd = sd(daysdeath, na.rm = TRUE))

cd_c8_subs %>% 
    mutate(
        death_by_1year = daysdeath <= 365 & !is.na(daysdeath)
    ) %>%
    count(death_by_1year)
cd_c8_subs %>% 
    mutate(
        death_by_2year = daysdeath <= 365*2 & !is.na(daysdeath)
    ) %>%
    count(death_by_2year)
```





# Summary statistics across covariates

For the participants in the mortality analysis:

```{r}
variables <- c("age", "gender", "race", "bmi", "cad", "cardiac_arrest", "cardiac_arrhythmias", "mi", "stroke", "cv_disease", "diabetes", "alb", "ktv", "dialysis_duration", "does_the_subject_have_clinically_documented_hypertension")
variable_types <- c("q", "c", "c", "q", rep("c", 7), rep("q", 3), "c")

show_summary_stats <- function(.data, vars, var_types) {
    for (i in seq_along(vars)) {
        var <- vars[i]
        cat(var, "-------------------------------------------:\n")
        if (var_types[i]=="c")
            .data %>% count(.data[[var]]) %>% mutate(frac = n/sum(n)) %>% print()
        else {
            print(summary(.data[[var]]))
        }
    }
}
show_summary_stats(cd_c8_subs, vars = variables, var_types = variable_types)
```

For LUCID participants excluded from mortality analysis:

```{r}
cd_c8 <- colData(se_c8_main) %>% as.data.frame() %>% filter(event_name=="Baseline")
bool_mort <- (cd_c8$country=="Canada" | cd_c8$frenova) & cd_c8$event_name == "Baseline"
bool_other <- !bool_mort

cd_c8_other <- cd_c8[bool_other,]
show_summary_stats(cd_c8_other, vars = variables, var_types = variable_types)
```

For all LUCID participants:

```{r}
show_summary_stats(cd_c8, vars = variables, var_types = variable_types)
```

Checking units for albumin in Canada and US. It does seem like Canada is in g/L and US in g/dL.

```{r}
print("alb")
summary(cd_c8_subs %>% filter(country=="Canada") %>% pull(alb))
summary(cd_c8_subs %>% filter(country=="US") %>% pull(alb))

print("hgb")
summary(cd_c8_subs %>% filter(country=="Canada") %>% pull(hgb))
summary(cd_c8_subs %>% filter(country=="US") %>% pull(hgb))

print("ferritin")
summary(cd_c8_subs %>% filter(country=="Canada") %>% pull(ferritin))
summary(cd_c8_subs %>% filter(country=="US") %>% pull(ferritin))

print("phosph")
summary(cd_c8_subs %>% filter(country=="Canada") %>% pull(phosph))
summary(cd_c8_subs %>% filter(country=="US") %>% pull(phosph))

print("ipth")
summary(cd_c8_subs %>% filter(country=="Canada") %>% pull(ipth))
summary(cd_c8_subs %>% filter(country=="US") %>% pull(ipth))

print("ktv")
summary(cd_c8_subs %>% filter(country=="Canada") %>% pull(ktv))
summary(cd_c8_subs %>% filter(country=="US") %>% pull(ktv))
```

Lab distributions for several lab measurements in Canada:

```{r}
summ_stats <- function(x) {
    cat("  Min:", min(x, na.rm = TRUE) %>% round(2), "\n")
    cat("  Max:", max(x, na.rm = TRUE) %>% round(2), "\n")
    cat("  Mean:", mean(x, na.rm = TRUE) %>% round(2), "\n")
    cat("  Median:", median(x, na.rm = TRUE) %>% round(2), "\n")
    cat("  SD:", sd(x, na.rm = TRUE) %>% round(2), "\n")
}
cat("Albumin:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(alb))

cat("Calcium:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(CA))

cat("Creatinine:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(CRE))

cat("Ferritin:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(ferritin))

cat("Hemoglobin:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(hgb))

cat("Iron:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(IRON))

cat("Phosphate:\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(phosph))

cat("Parathyroid hormone (CA):\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(ipth))

cat("Toral iron binding capacity (CA):\n")
summ_stats(cd_c8_subs %>% filter(country=="Canada") %>% pull(TIBC))
```

Sample sizes

```{r}
cat("Original sample size:", nrow(cd_c8_subs), "\n")

cd_c8_subs_nonlabs <- cd_c8_subs %>% select(age, gender, race, bmi, cv_disease, diabetes, dialysis_duration)
cc_nonlabs <- complete.cases(cd_c8_subs_nonlabs)
cat("# dropped due to missing data on race, bmi, CV disease, dialysis duration:", sum(!cc_nonlabs), "\n")

cd_c8_subs_all_covars <- cd_c8_subs %>% select(age, gender, race, bmi, cv_disease, diabetes, dialysis_duration, alb, ktv)
cc_all <- complete.cases(cd_c8_subs_all_covars)
cat("# dropped due to missing data across all covariates:", sum(!cc_all), "\n")
```


```{r}
cd_c8_subs %>% 
    filter(cc_nonlabs) %>%
    mutate(
        death_by_1year = daysdeath <= 365 & !is.na(daysdeath)
    ) %>%
    count(death_by_1year)
cd_c8_subs %>% 
    filter(cc_nonlabs) %>%
    mutate(
        death_by_2year = daysdeath <= 365*2 & !is.na(daysdeath)
    ) %>%
    count(death_by_2year)
```




# Filtering metabolites based on missingness and zero variance

This filtering was done earlier based on all 1000 samples. Are there any metabolites that remain 

- Remove metabolites with >= 95% missingness
- Remove metabolites where the variance in abundance among non-missing values is zero

```{r}
bool_c8 <- colData(se_c8_main) %>%
    as.data.frame() %>%
    mutate(keep = (country=="Canada" | frenova) & event_name == "Baseline") %>%
    pull(keep)
bool_hilic_pos <- colData(se_hilic_pos_main) %>%
    as.data.frame() %>%
    mutate(keep = (country=="Canada" | frenova) & event_name == "Baseline") %>%
    pull(keep)
bool_hilic_neg <- colData(se_hilic_neg_main) %>%
    as.data.frame() %>%
    mutate(keep = (country=="Canada" | frenova) & event_name == "Baseline") %>%
    pull(keep)
se_c8_main <- filter_metabs(se_c8_main[,bool_c8])
se_hilic_pos_main <- filter_metabs(se_hilic_pos_main[,bool_hilic_pos])
se_hilic_neg_main <- filter_metabs(se_hilic_neg_main[,bool_hilic_neg])
```





# `limma` analysis

Count number of metabolites.

```{r}
se_list <- list(
    c8 = se_c8_main,
    hilic_pos = se_hilic_pos_main,
    hilic_neg = se_hilic_neg_main
)

se <- merge_ses_across_injections(se_list)
nrow(se)
rowData(se) %>% as.data.frame() %>% count(is.na(hmdb_id))
```

`limma` analysis for 1-year and 2-year mortality.

- Full set of covariates: `alb` needs to be manually log-transformed; `ktv` is not log-transformed. `alb_imp` is already log-transformed.

```{r}
df_analysis_versions <- crossing(
    mod_form = c(
        ~mort_outcome+age+gender+race+cv_disease+diabetes+ns(bmi,df=3)+ns(log(alb),df=3)+ns(ktv,df=3)+ns(dialysis_duration,df=3)+country,
        ~mort_outcome+age+gender+race+cv_disease+diabetes+ns(bmi,df=3)+ns(alb_imp,df=3)+ns(ktv_imp,df=3)+ns(dialysis_duration,df=3)+country,
        ~mort_outcome
    ),
    metab_type = c("qrilc_no_ruv", "qrilc_ruv")
) %>% 
    mutate(analysis_version = c("ruv_no_imp_no", "ruv_yes_imp_no", "ruv_no_imp_yes", "ruv_yes_imp_yes", "ruv_no_unadj", "ruv_yes_unadj"))

limma_mort_results_list <- lapply(seq_len(nrow(df_analysis_versions)), function(i) {
    res <- run_mort_analysis(
        se_list = se_list,
        mod_form = df_analysis_versions$mod_form[[i]],
        metab_type = df_analysis_versions$metab_type[i],
        drug_metabs = drug_metabs
    )
    bind_rows(res, .id = "outcome") %>%
        mutate(analysis_version = df_analysis_versions$analysis_version[i])
})
```

Create `data.frame` of all results (across all sensitivity analyses).

```{r}
all_mort_results <- bind_rows(limma_mort_results_list) %>%
    mutate(metab_id = ifelse(is.na(metabolite), compound_id, metabolite))

all_mort_results %>% filter(adj_pval < 0.1)
all_mort_results %>% filter(adj_pval < 0.1, metabolite=="quinolinate")
all_mort_results %>% filter(adj_pval < 0.1, metabolite=="mesaconate")
```

ML analyses - known compounds only

```{r}
df_analysis_versions <- crossing(
    adj_vars = list(
        c("age", "gender", "race", "cv_disease", "diabetes", "bmi", "alb", "ktv", "dialysis_duration", "country"),
        c("age", "gender", "race", "cv_disease", "diabetes", "bmi", "alb_imp", "ktv_imp", "dialysis_duration", "country"),
        c()
    ),
    metab_type = c("qrilc_no_ruv", "qrilc_ruv")
) %>%
    mutate(analysis_version = c("ruv_no_imp_no", "ruv_yes_imp_no", "ruv_no_imp_yes", "ruv_yes_imp_yes", "ruv_no_unadj", "ruv_yes_unadj"))

set.seed(355)
ml_mort_models_list <- lapply(seq_len(nrow(df_analysis_versions)), function(i) {
    fit_ml_models_mort(
        se_list = se_list,
        adjustment_vars = df_analysis_versions$adj_vars[[i]],
        only_known_metabs = TRUE,
        metab_type = df_analysis_versions$metab_type[i]
    )
})
ml_mort_results_list <- lapply(seq_along(ml_mort_models_list), function(i) {
    res <- ml_mort_models_list[[i]]
    res_df <- ml_results_to_df(list(res))
    res_df %>%
        rename(outcome = symptom, importance_lasso = importance_lasso_1, importance_rf = importance_rf_1) %>%
        mutate(analysis_version = df_analysis_versions$analysis_version[i])
})

```

Report accuracy measures for ML models

```{r}
df_ml_acc <- lapply(seq_along(ml_mort_models_list), function(i) {
    mod_lasso <- ml_mort_models_list[[i]][["1_year_mort"]][[1]]
    mod_rf <- ml_mort_models_list[[i]][["1_year_mort"]][[2]]
    
    lasso_res <- mod_lasso$results %>% 
        filter(lambda==mod_lasso$bestTune$lambda)
    rf_res <- mod_rf$results %>% 
        filter(mtry==mod_rf$bestTune$mtry)
    
    tibble(
        analysis_version = df_analysis_versions$analysis_version[i],
        lasso_acc = lasso_res$Accuracy,
        lasso_acc_sd = lasso_res$AccuracySD,
        rf_acc = rf_res$Accuracy
    )
}) %>% bind_rows()
```

**Note:** ML analyses on the unknowns (`only_known_metabs = FALSE`) causes memory errors--too many unknowns

Merge ML results and limma results.

```{r}
df_ml_mort_results <- bind_rows(ml_mort_results_list)

all_mort_results_limma_ml <- all_mort_results %>% left_join(df_ml_mort_results, by = c("outcome", "metab_id", "analysis_version"))
```

Subset to known metabolites with significant associations.

```{r}
signif_metabs <- all_mort_results_limma_ml %>%
    filter(adj_pval < 0.1, !is.na(metabolite)) %>%
    pull(metabolite) %>%
    unique()
all_mort_results_limma_ml <- all_mort_results_limma_ml %>%
    mutate(FC = round(2^logFC, 2)) %>%
    select(outcome, method, metabolite, hmdb_id, logFC, FC, P.Value, adj_pval, importance_lasso, importance_rf, analysis_version, starts_with("mod_coeff_"))
all_mort_results_limma_ml_signif <- all_mort_results_limma_ml %>%
    filter(metabolite %in% signif_metabs) %>%
    arrange(metabolite, outcome)
```

```{r}
all_mort_results_limma_ml_signif %>% filter(metabolite %in% c("mesaconate", "quinolinate"), outcome=="1_year_mort") %>% select(metabolite, importance_lasso:analysis_version)
```

Write results to file.

```{r}
write_csv(df_ml_acc, file = "../results/tables/mort_ml_acc.csv")

write_csv(all_mort_results_limma_ml, file = "../results/tables/mort_results_all.csv")
write_csv(all_mort_results_limma_ml_signif, file = "../results/tables/mort_results_signif.csv")
```

